---
title: "Lee Carter Model"
output: pdf_document
---

```{r setup, echo=FALSE}
library(dplyr)
library(ggplot2)
library(demography)
library(readxl)
library(forecast)
library(StMoMo)
library(gridExtra)
```

##Describe Data
###Data Loading
  First we import the data for males, females and total population of Belgium and have a look at the variable names.
```{r, message=FALSE,error=FALSE}
female_Belgium <- read_excel("life tables_Belgium.xlsx", sheet = "females")
male_Belgium <- read_excel("life tables_Belgium.xlsx", sheet = "males")
all_Belgium <- read_excel("life tables_Belgium.xlsx", sheet = "total")

names(all_Belgium)
attach(female_Belgium)
```
###Data Inspection
  We first inspect the ranges of `Year` and `Age` for the three datasets, taking the female group as an example.
```{r}
summary(female_Belgium)
#summary(male_Belgium)
#summary(all_Belgium)
```
  
```{r}
cat("Female age ranges ", "from", min(female_Belgium$Age),
    "to",max(female_Belgium$Age),"during",min(female_Belgium$Year),"and",
    max(female_Belgium$Year),"\n")
cat("Male age ranges ", "from", min(male_Belgium$Age),
    "to",max(male_Belgium$Age),"during",min(male_Belgium$Year),"and",
    max(male_Belgium$Year),"\n")
cat("All age ranges ", "from", min(all_Belgium$Age),
    "to",max(all_Belgium$Age),"during",min(all_Belgium$Year),"and",
    max(all_Belgium$Year),"\n")

```
  It seems that the variable `Age` is recognized as a character vector, which is why the maximum age the function can recognize is 99, while the maximum age is actually 112.
  
###Descriptive Analysis
  Here are the graphs of the predictors against the response variable `log(mx)` for all groups. It shows that from 1914-1918 the central death rate data was missing. Therefore we decide to keep the data from 1919 on to 2018.
```{r message=FALSE,error=FALSE}
plot1<-ggplot(male_Belgium,aes(x= Year, y = log(qx)))+geom_point()+
  geom_line(stat='identity')+ggtitle("Beigium - males, 1840-2018")+labs(y = bquote(ln(q[x])))
plot2 <-ggplot(female_Belgium,aes(x= Year, y = log(qx)))+geom_point()+
  geom_line(stat='identity')+ggtitle("Beigium - females, 1840-2018")+labs(y = bquote(ln(q[x])))
plot3<- ggplot(all_Belgium,aes(x= Year, y = log(qx)))+geom_point()+
  geom_line(stat='identity')+ggtitle("Beigium - all, 1840-2018")+labs(y = bquote(ln(q[x])))
grid.arrange(plot1,plot2,plot3, ncol = 2)
```

```{r}
male <- male_Belgium%>%filter(Year >=1919)
female <- female_Belgium%>%filter(Year >=1919)
total <- all_Belgium%>% filter(Year >=1919)
```
  Afterwards, we inspect the mortality statistis in each group. To do this we define a function that plots the log central death rate against variables of interest. Here, after inspecting the first two graphs, we discover a strange pattern emerged for people aged between 20 and 30 or so. Apart from that, we can detect a linear trend between the logorithm central death rate and the variable of interest. Therefore, we remove the data for people aged less than 20. Here we assign new datasets to new names.
```{r}
#check statistics
checkstatistics <- function(male){
  k1<-ggplot(male, aes(Year,y = log(mx),color = Age))+geom_point()+geom_line(stat = 'identity')+
    theme(legend.position = "none")+ggtitle(paste("Belgium",gender," 1919-2018"))+ylab("Logrithmn of central death rate")
  #above: indicating a linear year effect on log(mx) for all ages 
  k2<-ggplot(male, aes(Age,y = log(mx),color = Year))+geom_point()+geom_line(aes(x= Age))+
    theme(legend.position = "none")+ggtitle(paste("Belgium",gender," 1919-2018"))+ylab("Logrithmn of central death rate")
  # above: indicating a different year effect on log(mx) in all ages 
  # for a group of a children, the death rate is significantly higher, so we decide to leave them out
  new_data <- male%>%mutate(logmx = log(mx))%>%filter(Age>20|logmx < -2.5 )
  k3 <-ggplot(new_data, aes(Age,y = log(mx),color = Year))+geom_point()+geom_line(aes(x= Age))+
    theme(legend.position = "none")+ggtitle(paste("Belgium",gender," 1919-2018"," aged 20-110"))+ylab("Logrithmn of central death rate")
  grid.arrange(k1,k2,k3,ncol = 2)
  return(new_data)
  }

male_new <- checkstatistics(male)
female_new<- checkstatistics(female)
all_new <- checkstatistics(total)
```
 To further affirm the linear trend, which consolidates our choice of using iterative least square method in our Lee Carter model, we substract one-year data in the male group as a demo. It shows that a linear trend sustains.
```{r, message=FALSE,error=FALSE}
#Choose 2018 as an example
male_2018 <- male_new %>% filter(Year ==2018)
ggplot(male_2018,aes(Age, y = log(mx)))+geom_point()+geom_line()

```

##Modelinng
###Model Assumptions
  *The model follows the following statistical expression:

$$
 log(m_{x,t}) = \alpha_x + \beta_{x,t}\times \kappa_{t} + e_{x,t} \\ where \ \alpha_x is\ the \ age \ effect, \\ \beta_{x,t} \ is \ the \ sensitivity  \ to \ the \ year \ effect \ \kappa_t .
$$
   *There exist a linear trend between the `Year` variable and the logorithm central death rate, and as a result the LS approach is adapted.
   
###Design Functions
  To estimate those parameters, let's first define some functions that can be applied to three groups.
```{r}
#find_alpha  : to return a plot of alpha values and the list of alpha values
find_alpha <- function(new){
  X <- model.matrix(
    ~ as.factor(new$Age) 
    - 1)
  dim(X)
  min(new$Age)
  max(new$Age)
  y <- log(new$mx)
  agemodel <- lm(y ~ -1+as.factor(new$Age))
  a_x <- agemodel$coefficients
  a_list <- data.frame(Age =agemodel$xlevels$`as.factor(new$Age)`,alpha = a_x)
  #visualize a list
  alphaplot <- ggplot(a_list,aes(x = Age, y = alpha))+geom_smooth(se=FALSE)+geom_point()+
    geom_line(aes(Age,alpha),stat = "identity")+
    ggtitle(paste("Belgium",gender, "1919-2018"))+ylab("Alpha")+theme_bw()
  ls <- list("plot" = alphaplot,"alist"=a_list)
  return(ls)
}
#find_kappa : return a plot of kappa values and big dataset with alpha, kappa values, and a single kappa values' dataset
find_kappa <-function(new,a_list){
  #estimate year effect
  #new variable z = log(mx)-a
  with_a <- left_join(new,a_list,by = "Age")
  z <- with_a$logmx - with_a$alpha
  yearmodel <- lm(z~-1+as.factor(new$Year))
  kappa_list <- data.frame(Year = yearmodel$xlevels$`as.factor(new$Year)`,kappa = yearmodel$coefficients)
  #visualize kappa
  kappaplot<- ggplot(kappa_list,aes(x = Year, y = kappa))+geom_point()+geom_smooth(se=FALSE)+
    theme_bw()+ggtitle(paste("Belgium",gender, "1919-2018"," starting values"))
  #combine kappa
  with_a$Year <- as.factor(with_a$Year)
  with_kappa <- left_join(with_a,kappa_list,by = "Year")
  ls <- list("plot"=kappaplot,"with_kappa"=with_kappa,"kappa_list"= kappa_list,"Z" = z)
  return(ls)
}
#find_beta : return a plot of beta values and big dataset with alpha,kappa and beta values, and a single beta values' dataset
find_beta <-function(with_kappa,z){
  betamodel <- lm(z~-1+as.factor(with_kappa$Age):with_kappa$kappa)
  beta_list <- data.frame(name = betamodel$xlevels$`as.factor(with_kappa$Age)`,
                          betas = betamodel$coefficients)
  #visualize betas
  beta_plot <- ggplot(beta_list,aes(x = name, y= betas))+geom_point()+geom_line()+
    ggtitle(paste("Belgium ",male," 1919-2018, starting values"))
  with_beta <- left_join(with_kappa,beta_list,by = c("Age" = "name"))
  ls <-list("plot"= beta_plot,"beta_list" = beta_list,"with_beta" = with_beta)
  return(ls)
}
#doconvergence : it is simply the convergence process and return the final converged beta/ kappa values
doconvergence <- function(beta_list,kappa_list,with_beta,z) {
  #convergence
  converged = F
  iter      = 1
  beta_est  = beta_list$betas
  kappa_est = kappa_list$kappa
  while(!converged){  
    beta_est_old  = beta_est
    kappa_est_old = kappa_est
    # (2): estimate kappa's
    var_beta = with_beta$betas
    X        = model.matrix(~ as.factor(with_beta$Year):var_beta - 1)
    kappa_est = solve(crossprod(X)) %*% t(X) %*% z
    
    # (3): estimate beta's
    var_kappa = with_beta$kappa
    X         = model.matrix(~ as.factor(with_beta$Age):var_kappa - 1)
    beta_est   = solve(crossprod(X)) %*% t(X) %*% z 
    
    # stopping criterion
    converged = 
      max(abs(beta_est - beta_est_old) / abs(beta_est_old), abs(kappa_est - kappa_est_old) / abs(kappa_est_old)) < 1e-8
    iter = iter + 1
    if(iter %% 1e2 == 0)
      cat("\n\nIteration number", iter, "\n\n")
 
  }
  ls <-list("beta_est"= beta_est,"kappa_est"= kappa_est)
  return(ls)
}
#visualize convergence: return final kappa and beta values for the final ploting and comparison
visualize_convergence <- function(kappa_list,beta_list,kappa_est,beta_est){
  kappafinal <- cbind(kappa_list,kappa_est)
  betafinal <- cbind(beta_list,beta_est)
  ggplot(kappafinal,aes(x=as.numeric(Year)))+geom_line(aes(y = kappa))+
    geom_line(aes(y = kappa_est),color="red")+xlab("Years")+ggtitle("Kappa final values V.S. starting values")
  ggplot(betafinal,aes(x = as.numeric(name)))+geom_point(aes(y = betas),color="black")+
    geom_line(aes(y = beta_est),color = "red")+xlab("Age")+ggtitle("Beta final values V.S. starting values")
  #apply constraints
  #sum(kappa_{x,t} = 0)
  #sum(beta_{x,t} = 1)
  ls <- list("kappafinal" = kappafinal, "betafinal" = betafinal)
  return(ls)
}
#find_ls: return the least-square estimates
find_ls <- function(beta_est,kappa_est,a_list,gender){
  beta_est <- apply(beta_est,2,as.numeric)
  kappa_est <- apply(kappa_est,2,as.numeric)
  alpha_ls <-apply(a_list,2,as.numeric)
  beta_ls <-beta_est/sum(beta_est)
  kappa_ls <- (kappa_est-mean(kappa_est))*sum(beta_est)
  alpha_ls <-a_list$alpha + beta_est*mean(kappa_est)
  cat(paste("The sum of beta_ls ", "for ",gender,"is ",sum(beta_ls),"\n"))
  cat(paste("The sum of kappa_ls ", "for",gender,"is",sum(kappa_ls),"\n"))
  ls <- list("bls" = beta_ls,"kls" = kappa_ls,"als"=alpha_ls)
  return(ls)
}
#plotfinals: final plots of alpha, kappa and beta values
plotfinals <- function(kappafinal,betafinal,a_list){
  kappafinal$Year<- as.numeric(kappafinal$Year)
  betafinal$name <- as.numeric(betafinal$name)
  a_list$Age <- as.numeric(a_list$Age)
  g1 <-ggplot(kappafinal,aes(x = Year, y = kappa_est))+geom_point()+geom_line()+
    ggtitle(paste("Kappa values for Belgium",gender,"1919-2018"))
  g2 <-ggplot(betafinal,aes(x = name, y = beta_est))+geom_point()+geom_line()+
    ggtitle(paste("Beta values for Belgium",gender,"1919-2018"))+xlab("Age")
  g3 <- ggplot(a_list,aes(x = Age,y = alpha))+geom_point()+geom_line()+
    ggtitle(paste("Kappa values for Belgium",gender,"1919-2018"))
  grid.arrange(g1,g2,g3, ncol=2)
}

```
###Apply Functions
  Now we apply those functions to our datasets and acquire the final estimates..
  
```{r}
gender = "male"
malpha_model <- find_alpha(new = male_new)
malpha_model$plot
mkappa_model <- find_kappa(new = male_new,a_list = malpha_model$alist)
mbeta_model <- find_beta(with_kappa = mkappa_model$with_kappa,z = mkappa_model$Z)
mconverged_est <- doconvergence(mkappa_model$kappa_list,
                               beta_list = mbeta_model$beta_list,
                               with_beta = mbeta_model$with_beta,
                               z=mkappa_model$Z)
mfinalest<- visualize_convergence(kappa_list = mkappa_model$kappa_list,
                                 beta_list = mbeta_model$beta_list,
                                 kappa_est = mconverged_est$kappa_est,
                                 beta_est = mconverged_est$beta_est)

#for female
gender = "female"
falpha_model <- find_alpha(new = female_new)
falpha_model$plot
fkappa_model <- find_kappa(new = female_new,a_list = falpha_model$alist)
fbeta_model <- find_beta(with_kappa = fkappa_model$with_kappa,
                        z = fkappa_model$Z)
fconverged_est <- doconvergence(fkappa_model$kappa_list,
                               beta_list = fbeta_model$beta_list,
                               with_beta = fbeta_model$with_beta,
                               z = fkappa_model$Z)
ffinalest<- visualize_convergence(kappa_list = fkappa_model$kappa_list,beta_list = beta_model$beta_list,
                                 kappa_est = fconverged_est$kappa_est,beta_est = converged_est$beta_est)
#for all populations
gender = "all"
lalpha_model <- find_alpha(new = all_new)
lalpha_model$plot
lkappa_model <- find_kappa(new = all_new,a_list = lalpha_model$alist)
lbeta_model <- find_beta(with_kappa = lkappa_model$with_kappa,
                        z = lkappa_model$Z)
lconverged_est <- doconvergence(lkappa_model$kappa_list,
                               beta_list = lbeta_model$beta_list,
                               with_beta = lbeta_model$with_beta,
                               z = lkappa_model$Z)
lfinalest<- visualize_convergence(kappa_list = lkappa_model$kappa_list,beta_list = lbeta_model$beta_list,
                                 kappa_est = lconverged_est$kappa_est,beta_est = lconverged_est$beta_est)


```

```{r}
#for males
male_ls <- find_ls(beta_est = mfinalest$betafinal,
                   kappa_est = mfinalest$kappafinal,
                   a_list = malpha_model$alist,gender = "male")
plotfinals(kappafinal = mfinalest$kappafinal,
           betafinal = mfinalest$betafinal,
           a_list = malpha_model$alist)
#for females
female_ls <- find_ls(beta_est = finalest$betafinal,
                     kappa_est = ffinalest$kappafinal,
                     a_list = falpha_model$alist,gender = "female")
plotfinals(kappafinal = ffinalest$kappafinal,betafinal = ffinalest$betafinal,a_list = falpha_model$alist)
#for all
all_ls <- find_ls(beta_est = lfinalest$betafinal,kappa_est = lfinalest$kappafinal,
                  a_list = lalpha_model$alist,gender = "all population")
plotfinals(kappafinal = lfinalest$kappafinal,betafinal = lfinalest$betafinal,a_list = lalpha_model$alist)
```

##Model Comparison
  To appraise the above models, we use the MSE statistics of the three models to examine the model's goodness of fit. First, we  used the estimates to compute predicted `mx` values and take a look at the QQ-plots.
```{R}
#A function to combine the data
finalmodel <- function(with_kappa, kappafinal,betafinal){
  with_final <- left_join(with_kappa,kappafinal,by = "Year" )
  with_final <- left_join(with_final,betafinal,by = c("Age" = "name"))
  with_final_2 <- with_final %>% mutate(newmx =exp(alpha + kappa_est*beta_est) )
  return(with_final_2)
}
#For three groups
mdata <- finalmodel(mkappa_model$with_kappa,mfinalest$kappafinal,mfinalest$betafinal)
ggplot(mdata,aes(x = mx,y = newmx))+geom_point()+
  geom_abline(intercept = 0,slope=1,size=5,color = "red")+
  ggtitle("QQ-plot for male")
fdata <- finalmodel(fkappa_model$with_kappa,ffinalest$kappafinal,ffinalest$betafinal)
ggplot(fdata,aes(x = mx,y = newmx))+
  geom_point()+
  geom_abline(intercept = 0,slope=1,size=5,color = "red")+
  ggtitle("QQ-plot for female")
ldata <- finalmodel(lkappa_model$with_kappa,lfinalest$kappafinal,lfinalest$betafinal)
ggplot(ldata,aes(x = mx,y = newmx))+
  geom_point()+
  geom_abline(intercept = 0,slope=1,size=5,color = "red")+
  ggtitle("QQ-plot for male")

```
  QQ plots all demonstrate pretty high predictivity, so we then turn to the MSE values of those models.
```{R}
cat("The MSE for males model is ",mean((mdata$mx-mdata$newmx)^2),"\n")
cat("The MSE for females model is ", mean((fdata$mx-fdata$newmx)^2),"\n")
cat("The MSE for total model is ", mean((ldata$mx-ldata$newmx)^2),"\n")

```
 Thus, the MSE value for the total model is slightly smaller than the two separate models. From this perspective we may suggest strategy (3) be slightly better, though they all show high accuracy.
 The joint model include all females and males in Belgium, and while people may question the gender effect, we may discover from the previous graphs that men and women seem to share the same age-death and year-death pattern. Thus, we do not consider the gender effect in the model.  
 
 The End.
